---
title: "NYC Flights"
author: "Ellen Creed, Rachel McInerney, Siún Mulcahy, Conor Thompson"
date: "`r format(Sys.time(), '%X %d %B, %Y')`"
output:
  html_document:
    code_folding: hide
---
 
## Introduction

This data describes all flights that departed New York City in 2013 and is available in R (R Core Team, 2023). 'flights' contains 327,346 observations and 19 variables detailing flight punctuality. Additional information regarding airlines, airports, weather, and
planes is also provided (Wickham, 2021).

Motivating questions include:  
1. Is there a particular airline that is more on time than others?  
2. What density distribution does flight duration follow?  
3. What patterns are seen among destination airports regarding origin airports?  
4. How do total flight trends vary throughout the year?

## Methods

To produce enlightening visualisations of this dataset, data cleaning and wrangling were performed with dplyr (Wickham et al., 2023). We used functions from own `densr` package (McInerney et al., 2024) to create a density histogram. Other plots were generated using ggplot2 (Wickham, 2016) and plotly was used for interactive graphics (Sievert, 2020).

## Results

```{r}
# Install packages:
# install.packages('devtools') 
# install.packages('nycflights13')
# install.packages("usmap")
# install.packages("mapdata")
# install.packages("fpp3")
# install.packages("plotly")
# install.packages("RColorBrewer")
# devtools::install_github('rachel-mc/densr')

# Load the relevant libraries
suppressMessages(library(nycflights13))
suppressMessages(library(densr))
suppressMessages(library(tidyverse))
suppressMessages(library(ggplot2))
suppressMessages(library(fpp3))
suppressMessages(library(usmap))
suppressMessages(library(mapdata))
suppressMessages(library(plotly))
suppressMessages(library(RColorBrewer))

# Remove NAs
flights <- nycflights13::flights[complete.cases(nycflights13::flights[,]),] 
```


### Question 1

Adopted approach: an airline is the most on time if it has the smallest arrival delay

```{r}
average_delay <- flights |>
  group_by(carrier) |>
  summarise(avg_arrival_delay = mean(arr_delay))

ggplot(flights, aes(x = carrier, y = arr_delay)) +
  geom_boxplot(fill = "skyblue", colour = "blue") +
  labs(title = "Arrival Delay by Airline",
       x = "Airline",
       y = "Arrival Delay (minutes)") +
  theme_minimal() +
  ylim(c(-75, 100)) + # to zoom in and account for outliers
  # negative arrival delay corresponds to time gained during a journey i.e. passengers arrive earlier than scheduled
  coord_flip() 
```

This horizontal boxplot representing the average arrival delay in minutes of different carriers in the data displays that 'AS' (Alaska Airlines) exhibits the least arrival delays, depicted by the position of the central tendency. This is closely followed by 'HA' (Hawaiian Airlines Inc.), so these airlines are more on time than others. In stark contrast, the carriers 'FL' and 'F9' display the longest arrival delay, meaning that they are the most delayed airlines. Their outliers are positioned on the positive side of the x-axis, implying that many observations are more delayed than the average. Notably, ‘AS’ and ‘HA’ have a relatively large spread, especially in comparison to 'OO' (SkyWest Airlines Inc.) whose arrival delay times are represented by a particularly small boxplot. Therefore, even though 'AS' is the most on time on average, there is some variability within its values. 'OO' demonstrates the most consistent performance especially since it also has the least number of outliers. If we focus solely on the boxplots’ interquartile ranges and means, this plot highlights 'AS' and 'HA' as favourable choices for passengers prioritising punctuality while 'FL' and 'F9' are poor choices as they perform weakest in terms of arrival delay.


### Question 2

```{r}
# Initial data manipulation
# Group the data into individual planes and compute the median air time (more robust)
air_num <- flights |>
  group_by(tailnum) |>
  summarise(airtime = median(air_time))

# Extract the numeric vector converted to hours
air_time <- air_num$airtime/60

# Overlay frequency polygon density estimation
frequency_polygon(air_time)

air_fp <- create_hist_dens(air_time, type = "fp", breaks = "Scott")

suppressWarnings(plot(air_fp, border = "dodgerblue2", lwd = 1.5, lty = 2, 
                      xlab = "Median Air Time in hours per Airline", 
                      main = "Histogram of air_time with Frequency Polygon Density Estimate")) 
```

The objective here was to find the distributional form for the density of observed ‘air_time’ using a continuous histogram estimator. This density distribution appears to be bimodal, with the frequency polygon coordinates capturing the second significant peak. The distribution is right-skewed, and we observe positive kurtosis. The most frequent median flight time per plane lasted just over two hours. There are outliers appearing at flight durations of over 10 hours. The plane that spent the most time in the air in 2013 on average was N384HA (see Suppl. Info.). Consulting the planes metadata highlighted that this was a 2011 Fixed wing multi-engine manufactured by AIRBUS. Intuitively, the different routes between airports explain the varying lengths of airtime. Additional analysis could uncover which specific planes lie in the quantiles of this data, if external factors such as weather affect airtime, and so on.


### Question 3

```{r}
#Airports Data
airports <- nycflights13::airports

airport_flights <- flights %>% group_by(origin)

#Finding Unique destinations 
dest <- flights$dest
unique(dest)

##Long and Lats for Map
#Creating a dataframe with long and lats of each airport
airport_labels <- data.frame(
  airport = c("BWI", "CLT", "BUF", "DEN", "SNA", "MSY", "SLC", "XNA", "MKE", "SEA", "ROC", "SYR", "SRQ", "RDU", "CMH", "JAX", "CHS", "MEM", "PIT", "SAN", "DCA", "CLE", "STL", "MYR", "JAC", "MDW", "HNL", "BNA", "AUS", "BTV", "PHL", "STT","EGE", "AVL", "PWM", "IND", "SAV", "CAK", "HOU", "LGB", "DAY", "ALB", "BDL", "MHT", "MSN", "GSO", "CVG", "BUR", "RIC", "GSP", "GRR", "MCI", "ORF", "SAT", "SDF", "PDX", "SJC", "OMA", "CRW", "OAK", "SMF", "TUL", "TYS", "OKC", "PVD", "DSM", "PSE", "BHM", "CAE", "HDN", "BZN", "MTJ", "EYW", "PSP", "ACK", "BGR", "ABQ", "ILM", "MVY", "SBN", "LEX", "CHO", "TVC", "ANC", "LGA", "ATL", "BOS", "BQN", "DFW", "DTW", "FLL", "IAD", "IAH", "LAS", "LAX", "MCO", "MIA", "MSP", "ORD", "PBI", "PHX", "RSW", "SFO", "SJU", "TPA"),
  dest = c("BWI", "CLT", "BUF", "DEN", "SNA", "MSY", "SLC", "XNA", "MKE", "SEA", "ROC", "SYR", "SRQ", "RDU", "CMH", "JAX", "CHS", "MEM", "PIT", "SAN", "DCA", "CLE", "STL", "MYR", "JAC", "MDW", "HNL", "BNA", "AUS", "BTV", "PHL", "STT", "EGE", "AVL", "PWM", "IND", "SAV", "CAK", "HOU", "LGB", "DAY", "ALB", "BDL", "MHT", "MSN", "GSO", "CVG", "BUR", "RIC", "GSP", "GRR", "MCI", "ORF", "SAT", "SDF", "PDX", "SJC", "OMA", "CRW", "OAK", "SMF", "TUL", "TYS", "OKC", "PVD", "DSM", "PSE", "BHM","CAE", "HDN", "BZN", "MTJ", "EYW", "PSP", "ACK", "BGR", "ABQ", "ILM","MVY", "SBN", "LEX", "CHO", "TVC", "ANC", "LGA","ATL", "BOS", "BQN", "DFW", "DTW", "FLL", "IAD", "IAH", "LAS", "LAX", "MCO", "MIA", "MSP", "ORD", "PBI", "PHX", "RSW", "SFO", "SJU", "TPA"),
   lat = c(39.1754, 35.214, 42.940, 39.856, 33.6757, 29.9934, 40.7899, 36.2819, 42.9485, 47.4502, 43.1185, 43.1185, 27.9779, 35.8801, 39.998, 30.4941, 32.8987, 35.0423, 40.4958, 32.7336, 38.8512, 41.409, 38.7487, 33.6827, 43.607, 41.7868, 21.3241, 36.1249, 30.1944, 44.472, 39.872, 18.3371, 39.6426, 35.4362, 43.6445, 39.7173, 32.1316, 40.9168, 29.6454, 33.8171, 39.9023, 42.748, 41.9389, 42.9309, 43.1399, 36.0977, 39.0462, 34.2005, 37.5067, 34.8903, 42.8864, 39.2978, 36.8947, 29.5337, 38.178, 45.5887, 37.3645, 41.3031, 38.3721, 37.7214, 38.6953, 36.1984, 35.8117, 35.3931, 41.7265, 41.532, 18.0086, 33.5618, 33.9382, 40.4872, 45.7775, 38.5111, 24.5567, 33.8223, 41.2535, 44.8074, 35.0496, 34.2704, 41.3917, 41.7086, 38.0367, 38.1386, 44.7414, 61.1743, 40.7769, 33.6407, 42.3656, 18.4955, 32.8965, 42.2129, 26.0727, 38.9476, 29.9902, 36.084, 33.9416, 28.4312, 25.7933, 44.882, 41.9796, 26.6832, 33.4351, 26.5362, 37.6189, 18.4394, 27.979168),
  long = c(-76.6683, -80.9431, -78.732, -104.6737, -117.8678, -90.2572, -111.9791, -94.306, -87.8974, -122.3093, -77.6267, -76.1117, -82.5536, -78.787, -82.883, -81.6879, -80.0405, -89.9767, -80.2564, -117.1973, -77.0377, -81.8498, -90.370, -78.9288, -110.737, -87.7521, -157.925, -86.6702, -97.6699, -73.155, -75.2432, -64.9738, -106.9155, -82.5411, -70.3063, -86.2945, -81.2014, -81.4422, -95.2789, -118.1525, -84.2209, -73.8017, -72.6832, -71.437, -89.3375, -79.9373, -84.6623, -118.358, -77.3195, -82.2179, -85.5306, -94.7139, -76.202, -98.4691, -85.7396, -122.5976, -121.928, -95.894, -81.593, -122.2208, -121.591, -95.8836, -83.985, -97.6009, -71.4391, -93.6486, -66.5634, -86.752, -81.1189, -107.2219, -111.153, -107.8936, -81.7595, -116.5076, -70.0602, -68.8284, -106.617, -77.9025,-70.6156, -86.3175, -84.6042, -78.4529, -85.5828, -149.9962, -73.8734, -84.4277, -71.0096, -67.1342, -97.038, -83.3534, -80.1528, -77.4558, -95.339, -115.152, -118.4085, -81.308, -80.2906, -93.2218, -87.882, -80.0956, -112.0068, -81.7555, -122.375, -66.0018, -82.539337))

#Data Manipulation
#Number of flights from each origin airport to each destination airport:
flight_count_per_airport <- flights |>
  group_by(dest, origin) |>
  summarise(flight_count_per_airport = n(), .groups = "drop")

flight_count_per_airport <- flight_count_per_airport |>
  group_by(dest)

#Total number of flights in each destination:
count <- flights |> 
  select(dest) |>
  count(dest)

count <- as.data.frame(count)

count <- count |> rename(Total_Flight_Count = n)

#Merging data sets:
airportCount <- merge(airport_labels, count, by.x = "airport", by.y = "dest", all.x = TRUE)

all_combinations <- expand_grid(dest = unique(flights$dest), origin = c("JFK", "LGA", "EWR"))

# Group by dest and origin and calculate flight counts
flight_count_per_airport <- flights %>%
  count(dest, origin) %>%
  right_join(all_combinations, by = c("dest", "origin")) %>%
  mutate(flight_count = replace_na(n, 0)) %>%
  select(-n)

# Pivot the data to have separate columns for each origin
flight_count_per_airport <- flight_count_per_airport %>%
  pivot_wider(names_from = origin, values_from = flight_count, values_fill = 0)

# Calculate total flight count per destination
total_flight_count <- flight_count_per_airport %>%
  group_by(dest) %>%
  summarise(total_flight_count = sum(JFK, LGA, EWR), .groups = "drop")

# Join with total_flight_count
flight_count_per_airport <- flight_count_per_airport %>%
  left_join(total_flight_count, by = "dest")

# Join with airport_labels
flight_count_per_airport <- flight_count_per_airport %>%
  left_join(airport_labels, by = "dest")

#Labels for plot
flight_count_per_airport$text <- paste("Airport:", flight_count_per_airport$airport,
                                       "<br>Total Flight Count:", flight_count_per_airport$total_flight_count,
                                       "<br>Flights from JFK:", flight_count_per_airport$JFK,
                                       "<br>Flights from LGA:", flight_count_per_airport$LGA,
                                       "<br>Flights from EWR:", flight_count_per_airport$EWR)

flight_count_per_airport

#Counting airport with highest count per dest 
flight_count_per_airport <- flight_count_per_airport %>%
  mutate(highest_origin = case_when(
    JFK > EWR & JFK > LGA ~ "JFK",
    EWR > JFK & EWR > LGA ~ "EWR",
    LGA > JFK & LGA > EWR ~ "LGA",
    TRUE ~ NA_character_
  ))

# Plot
usa_map <- map_data("usa")

mapUSA <- ggplot() +
  geom_polygon(data = usa_map, aes(x = long, y = lat, group = group), 
               fill = "lightgrey") +
  geom_path(data = map_data("state"), aes(x = long, y = lat, group = group), size = 0.5) +
  geom_point(data = flight_count_per_airport, aes(x = long, y = lat, size = total_flight_count, fill = highest_origin, text = text), 
             shape = 21, color = "grey") +  
  scale_size_continuous(name = "Total Flight Count") +
  scale_color_manual(name = "Highest Origin",
                     values = c("JFK" = "green", "EWR" = "red", "LGA" = "blue")) +  
  coord_fixed() +
  theme_void()+
  ggtitle("Flight Counts from NYC Airports to Airports in the USA")

mapInteractive <- ggplotly(mapUSA, tooltip = c("text"))

mapInteractive

htmlwidgets::saveWidget(widget = mapInteractive, file = "Q3Interactive.html")
```

This map visualizes the United States, with each point representing one of the NYC airports JFK, EWR, or LGA and the destination of each flight in 2013. The size of each point corresponds to the total flight count, providing a quick visual indicator of airport activity. Additionally, the colour of each point reflects the origin airport with the highest flight count, offering insights into the main gateway for each location. LEX in Kentucky stands out as the least connected airport, with only one flight originating from NYC. In contrast, ORD airport in Chicago emerges as the most popular destination, with 17,283 flights recorded that primarily originated from LGA. Interestingly, some states like North Dakota and South Dakota have airports but lack any points on the map. This absence could stem from no flights in 2013 or missing data. We also observe the role of each airport; western states tend to have a higher frequency of flights originating from JFK airport, likely due to its status as NYC's major international hub that serves longer-distance flights. Conversely, states closer to NYC predominantly use EWR and LGA airports, which are smaller for shorter-distance flights. 

### Question 4

```{r, message = FALSE, warning = FALSE}
#Data Manipulation
flights <- flights %>%
  mutate(day_of_year = as.numeric(format(time_hour, "%j")), carrier = as.factor(carrier)) %>% 
  mutate(Date = as.Date(time_hour))

flights_totals <- flights %>% 
  group_by(carrier, day_of_year) %>% 
  summarise(total_flights = n()) %>% 
  ungroup()

flights_sum <- flights_totals %>% 
  group_by(day_of_year) %>% 
  summarise(sum_total = sum(total_flights)) %>% 
  ungroup()

carrier_totals <- aggregate(total_flights ~ carrier , data = flights_totals, FUN = sum)

carrier_totals_ordered <- carrier_totals %>% 
  arrange(desc(total_flights))

top_4_carriers <- head(carrier_totals_ordered, 4)$carrier

flights_totals_Date <- flights %>% 
  group_by(Date, carrier) %>% 
  summarise(total_flights = n()) %>% 
  ungroup()

flights_totals_UA <- flights[flights$carrier == "UA",] %>% 
  group_by(Date) %>% 
  summarise(total_flights = n()) %>% 
  ungroup()
flights_totals_B6 <- flights[flights$carrier == "B6",] %>% 
  group_by(Date) %>% 
  summarise(total_flights = n()) %>% 
  ungroup()
flights_totals_DL <- flights[flights$carrier == "DL",] %>% 
  group_by(Date) %>% 
  summarise(total_flights = n()) %>% 
  ungroup()
flights_totals_EV <- flights[flights$carrier == "EV",] %>% 
  group_by(Date) %>% 
  summarise(total_flights = n()) %>% 
  ungroup()

dcmp_UA <- flights_totals_UA %>% 
  as_tsibble() %>% 
  model(stl = STL(total_flights))
dcmp_B6 <- flights_totals_B6 %>% 
  as_tsibble() %>% 
  model(stl = STL(total_flights))
dcmp_DL <- flights_totals_DL %>% 
  as_tsibble() %>% 
  model(stl = STL(total_flights))
dcmp_EV <- flights_totals_EV %>% 
  as_tsibble() %>% 
  model(stl = STL(total_flights))

comp_UA <- components(dcmp_UA) %>% 
  as_tsibble()
comp_B6 <- components(dcmp_B6) %>% 
  as_tsibble()
comp_DL <- components(dcmp_DL) %>% 
  as_tsibble()
comp_EV <- components(dcmp_EV) %>% 
  as_tsibble()

major_holidays <- as.Date(c("2013-07-04", "2013-09-01", "2013-11-28", "2013-12-25", "2014-01-01" ))

ggplot(subset(flights_totals_Date, carrier %in% top_4_carriers), aes(x = Date, y = total_flights, colour = carrier)) +
  geom_line(alpha = 0.6) +
  geom_line(data = comp_UA, aes( y = trend), colour = "purple") + 
  geom_line(data = comp_B6, aes( y = trend), colour = "red") +
  geom_line(data = comp_DL, aes( y = trend), colour = "darkgreen") +
  geom_line(data = comp_EV, aes( y = trend), colour = "blue") + 
  geom_vline(xintercept = major_holidays, lty = "dashed", alpha = 0.8) + 
  labs(y= "Total Flights", x = "Date", title = "Flight Trends across 4 Major Airlines in NYC")
```

This plot illustrates the total number of flights per day across three major airports in NYC: John F Kennedy Intl, La Guardia, and Newark Liberty Intl. These totals are associated with the four most popular airlines in the dataset. While we see a mostly stationary time series, there is notable weekly seasonality in which weekends experience lower activity compared to weekdays (see Suppl. Info.). Flight trendlines for each airline are superimposed on the actual data and offer insights into periods when airports experienced higher-than-usual activity. Evidently, there are distinct dips in flight count on specific dates. To investigate these dips, an examination of the weather dataset was conducted, but no anomalies were found for these days. A list comprising major federal holidays in the US was then created. Overlaying vertical lines that signify these dates indicated a correlation with some of the observed troughs. Additionally, carrier B6 appears to be slightly more resilient to these holiday-related trends. This could be attributed to the fact that B6's primary airport is JFK, which is not the primary airport for any of the other major airlines plotted. However, further analysis would be required to prove this hypothesis.

## Discussion

The visualisations above display this data coherently so that we can make statistical inferences from them. Approximating the distribution of a variable is a powerful initial step to uncover prevailing truths in data. Additionally, we received insights into which airlines tended to have late arrivals, which locations were often flown into, and times of the year when flight sales plummeted. Airline companies could use such information for future planning needs. 

## References

McInerney R, Thompson C, Ukachukwu O (2024). _densr: Overlay densities to
histograms_. R package version 0.1.0, <https://github.com/rachel-mc/densr>.  

R Core Team. (2023). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL: https://www.R-project.org/ 

C. Sievert. Interactive Web-Based Data Visualization with R, plotly, and shiny.
Chapman and Hall/CRC Florida, 2020.

Wickham H (2016). ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York. ISBN 978-3-319-24277-4, https://ggplot2.tidyverse.org.  

Wickham H (2021). _nycflights13: Flights that Departed NYC in 2013_. R package version 1.0.2, <https://CRAN.R-project.org/package=nycflights13>.  

Wickham H, François R, Henry L, Müller K, Vaughan D (2023). _dplyr: A Grammar of Data Manipulation_. R package version 1.1.4, <https://CRAN.R-project.org/package=dplyr>.
